---
title: "v22 Workflow Vignette: Spatial eDNA + qPCR Joint Model"
author: "Kelly Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.width  = 9,
  fig.height = 6
)
```

## Overview

This vignette demonstrates the v22 spatial joint model for eDNA metabarcoding
and qPCR data.  The model estimates spatially varying log-concentrations for
multiple species simultaneously, using:

- **eDNA reads**: Poisson zero-inflation → Beta-Binomial proportional model
- **qPCR**:       Per-replicate hurdle model (Bernoulli detection + Normal log-concentration)
- **Spatial field**: Isotropic 2D Hilbert-space Gaussian Process (HSGP) with a
  single length scale `rho[sp]` per species

**Key v22 feature**: single isotropic `rho[sp]` replaces the anisotropic
`rho_x[sp] + rho_y[sp]` pair from v21.  For ecologically isotropic fields this
eliminates a bimodal posterior (two equivalent length-scale solutions) and
simplifies prior specification.

**Spatial model**:
$$\log\lambda_{l,s} = \psi + \delta_s + f_s(\mathbf{x}_l) + \epsilon_{l,s}$$

where $\psi$ = population mean, $\delta_s$ = species offset, $f_s$ = isotropic
GP (HSGP approximation), and $\epsilon_{l,s}$ = iid nugget noise.

---

## 1. Setup

```{r load-workflow}
# Source the self-contained workflow file — no other dependencies needed.
# Adjust the path if running from a different working directory.
source("run_qpcr_metabar_workflow_v22.R")
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
rstan_options(threads_per_chain = 4)

```

The workflow file loads `tidyverse`, `rstan`, `sf`, `akima`, and `patchwork`
and defines all functions in a single file.

---

## 2. Simulate Example Data

We generate a synthetic survey with:

- 50 sampling locations in a ~400 × 600 km domain (Pacific Northwest coast)
- 4 species with isotropic spatial structure
- `rho_km = 80` km length scale (same for all species in this example)
- `gp_alpha` drawn randomly per species (~0.5–2.0)
- `log_lambda_mean = log(4)` → moderate eDNA density (~35% zeros)

```{r simulate}
sim <- simulate_qpcr_metabar_data_v22(
  n_locations                   = 50,
  n_species                     = 4,
  reference_species             = 1,
  n_qpcr_per_location           = 3,
  rho_km                        = 80,          # isotropic GP length scale (km)
  gp_alpha_true                 = c(1.2, 0.8, 1.5, 1.0),
  sigma_nugget_true             = 0.1,        # per-location noise (all species)
  log_lambda_mean               = log(20),      # mean log-copies/aliquot
  sigma_qpcr_true               = 0.3,         # qPCR measurement SD (log scale)
  qpcr_detection_intercept_true = -2.0,
  qpcr_detection_slope_true     = 1.5,
  beta0_true                    = 5.0,         # Beta-Binomial dispersion intercept
  beta1_true                    = -2.0,        # amplification effect on dispersion
  gamma0_true                   = 2.0,         # low-copy inflation magnitude
  gamma1_true                   = exp(1.0),    # low-copy inflation rate
  total_reads_per_rep           = 10000,
  seed                          = 17
)

cat("Species names:", paste(sim$species_names, collapse = ", "), "\n")
cat("Reference species:", sim$species_names[sim$reference_species], "\n")
```

### What the simulator returns

```{r sim-structure}
names(sim)
names(sim$true_values)
```

- `edna_data` — long-format data frame: one row per (PCR replicate, species)
- `qpcr_data` — one row per qPCR replicate of the reference species
- `locations`  — lat/lon for each sampling location
- `true_values` — ground truth for model validation (concentrations, GP fields, etc.)

### Quick look at the data

```{r sim-preview}
# eDNA data: reads per species per PCR replicate
head(sim$edna_data, 12)

# qPCR data: binary detection + log-concentration when detected
head(sim$qpcr_data[sim$qpcr_data$detected == 1, ], 6)

# Zero proportion in eDNA (Poisson-driven)
zero_rate <- mean(sim$edna_data$Nreads == 0)
cat(sprintf("eDNA zero rate: %.1f%%\n", 100 * zero_rate))
```

---

## 3. Format Data

`format_qpcr_metabar_data_v22` clusters nearby coordinates, converts to UTM
(centered), and constructs the Stan-ready arrays.  Any location where **all**
qPCR replicates are non-detected is automatically excluded (the hurdle qPCR
likelihood requires at least one detected replicate per location).

```{r format}
fmt <- format_qpcr_metabar_data_v22(
  edna_data         = sim$edna_data,
  qpcr_data         = sim$qpcr_data,
  alpha_known       = sim$alpha_known,
  reference_species = sim$reference_species,
  species_names     = sim$species_names,
  pcr_id_col        = "pcr_replicate_id"
)
```

Key fields in the returned list:

| Field | Description |
|---|---|
| `N_locations` | Number of included locations |
| `N_edna_obs` | Number of eDNA PCR replicates |
| `N_qpcr_reps` | Number of qPCR replicates |
| `edna_obs_data` | `[N_edna_obs × N_species]` read count matrix |
| `x_utm`, `y_utm` | Centered UTM coordinates (km) |
| `has_edna`, `has_qpcr` | Per-location availability flags |

```{r format-summary}
cat("Locations retained:", fmt$N_locations, "\n")
cat("UTM domain (km):", round(diff(range(fmt$location_list$x_utm)), 1), "x",
    round(diff(range(fmt$location_list$y_utm)), 1), "\n")
```

---

## 4. Build Stan Data

`make_stan_data_v22` sets HSGP domain parameters and prior hyperparameters.

**HSGP domain sizing** (`m_x`, `m_y`, `max_rho`):

- `m_x × m_y` basis functions approximate the 2D GP; higher = more accurate but slower
- `max_rho` sets the boundary half-width: `L = max_rho × π / 4`
  - Set `max_rho` ≥ largest plausible true rho
  - With `max_rho = 300`, the domain can represent length scales up to ~300 km

**Key prior choices**:

| Parameter | Prior | Notes |
|---|---|---|
| `rho` | Truncated-Normal(mean, sd) > 0 | Use broad SD to allow data to inform |
| `gp_alpha` | Log-Normal(mean, sd) | Mean ≈ 1 (log scale); adjust to expected amplitude |
| `sigma_nugget` | Log-Normal hier. | `prior_sd ≈ 1e-5` pins nugget ≈ exp(prior_mean) |
| `beta0/beta1/gamma0/gamma1` | Dispersion model priors | See Stan file for details |

```{r stan-data}
sdat <- make_stan_data_v22(
  formatted_data          = fmt,
  N_pcr                   = 35,       # total PCR cycles in the qPCR assay
  m_x                     = 3,       # HSGP basis functions, E-W
  m_y                     = 4,        # HSGP basis functions, N-S
  max_rho                 = 300,      # km; must exceed largest expected rho
  rho_prior_mean          = 100,      # truncated-normal center (km)
  rho_prior_sd            = 200,      # broad SD lets data identify rho
  gp_alpha_prior_mean     = 1.0,      # log-normal mean on log scale
  gp_alpha_prior_sd       = 0.5,
  sigma_nugget_prior_mean = log(0.1), # pins hierarchical nugget near 0.25
  sigma_nugget_prior_sd   = 1e-3,     # near-zero SD = fixed nugget (diagnostic mode)
  sigma_qpcr_prior_sd     = 1.0,
  beta0_prior_sd          = 1.0,
  gamma0_prior_mean       = 2.0,
  gamma0_prior_sd         = 1.0,
  log_gamma1_prior_mean   = 1.0,
  log_gamma1_prior_sd     = 0.75
)

cat("Stan data dimensions: N_locations =", sdat$N_locations,
    "| N_species =", sdat$N_species,
    "| M (basis) =", sdat$m_x * sdat$m_y, "\n")
```

---

## 5. Fit the Model

```{r fit, results='hide'}
res <- fit_and_plot_v22(
  formatted_data  = fmt,
  stan_file       = "quant_metabar_qpcr_v22.stan",
  stan_data       = sdat,
  chains          = 3,
  iter            = 1000,
  warmup          = 500,
  seed            = 42,
  adapt_delta     = 0.99,
  max_treedepth   = 14,
  true_values     = sim$true_values,   # pass TRUE values for validation plots
  save_outputs    = TRUE,
  output_dir      = "outputs_v22"
)
```

> **Runtime**: 3 chains × 2000 iterations at 50 locations typically takes
> 20–45 minutes.  For quick testing use `chains = 1, iter = 600, warmup = 300`.

### Or use the one-call shortcut

```{r run-workflow, eval=FALSE}
# Equivalent to steps 2–5 in a single call:
res <- run_workflow_v22(
  n_locations  = 60,
  n_species    = 4,
  rho_km       = 80,
  stan_file = "quant_metabar_qpcr_v22.stan",
  chains       = 3,
  iter         = 1000,
  warmup       = 500,
  m_x = 5,
  m_y = 6,
  rho_prior_mean = 100,
  rho_prior_sd   = 200,
  sigma_nugget_true        = 0.05,
  sigma_nugget_prior_mean  = log(0.1),
  sigma_nugget_prior_sd    = 1e-3,
  log_lambda_mean = log(30),
  seed = 73
)
```

---

## 6. Diagnostics

Check convergence before interpreting results.

```{r diagnostics}
diag <- res$diagnostics
cat(sprintf("Divergent transitions : %d\n",   diag$n_divergent))
cat(sprintf("Max treedepth hits    : %d\n",   diag$max_treedepth_hits))
cat(sprintf("Max Rhat              : %.4f\n", diag$max_rhat))
cat(sprintf("Parameters Rhat > 1.01: %d\n",  diag$n_high_rhat))

# Parameters with the worst mixing
print(diag$top_rhat)
```

**Interpreting diagnostics**:

- **Divergences > 0**: indicates difficult geometry; try higher `adapt_delta` (e.g. 0.995) or reparameterization
- **Max treedepth hits > 0**: increase `max_treedepth` (e.g. 15) or use fewer basis functions
- **Rhat > 1.01**: poor mixing; increase iterations or investigate prior-likelihood conflicts

---

## 7. Parameter Estimates

### Spatial length scale (rho)

```{r rho-summary}
res$rho_summary$species_summary
```

The `eff_range` column = `2.15 × rho`: the distance at which spatial correlation
drops to ~0.05 under the squared-exponential kernel.

```{r rho-plot, fig.height=3}
res$rho_posterior_plot
```

Red vertical lines = true values.  Posteriors should cover the true `rho_km = 80`.

### GP amplitude (gp_alpha)

```{r gp-alpha}
res$gp_alpha_summary$species_summary
```

```{r gp-alpha-plot, fig.height=4}
res$gp_alpha_posterior_plot
```

### Nugget variance

```{r nugget}
res$nugget_summary$species_summary
```

```{r var-decomp, fig.height=4}
res$var_decomp_plot
```

The stacked bar shows the fraction of log-concentration variance attributable to
the GP spatial field vs. the iid nugget per species.  A high nugget fraction
suggests locally variable concentrations that the smooth GP cannot capture.

### qPCR parameters

```{r qpcr-params}
# sigma_qpcr
cat(sprintf("sigma_qpcr : %.3f (95%% CI: %.3f-%.3f)\n",
            res$qpcr_params$sigma_qpcr$mean,
            res$qpcr_params$sigma_qpcr$q025,
            res$qpcr_params$sigma_qpcr$q975))

# Detection curve
cat(sprintf("intercept  : %.3f (%.3f-%.3f)\n",
            res$qpcr_params$qpcr_detection_intercept$mean,
            res$qpcr_params$qpcr_detection_intercept$q025,
            res$qpcr_params$qpcr_detection_intercept$q975))
cat(sprintf("slope      : %.3f (%.3f-%.3f)\n",
            res$qpcr_params$qpcr_detection_slope$mean,
            res$qpcr_params$qpcr_detection_slope$q025,
            res$qpcr_params$qpcr_detection_slope$q975))
```

---

## 8. Model Fit Plots

### qPCR fit

Observed vs. posterior-predicted log-concentration for detected replicates.
X marks = non-detected replicates (plotted at the posterior mean for reference).

```{r qpcr-fit}
res$qpcr_plot
```

### Detection probability curve

Detection probability as a function of posterior log-concentration at each
location.  Dots = observed detection rate; size = number of replicates.

```{r detection-curve}
res$detection_plot
```

### eDNA proportion fit

Observed vs. posterior-predicted mean proportion of reads for each species.
Points should cluster near the 1:1 line.

```{r edna-fit, fig.height=8}
res$combined_edna
```

---

## 9. Spatial Concentration Maps

Interpolated posterior-mean concentration surface per species.  Points = sampling
locations, colored by their true concentration (in simulations) or posterior mean
(real data).

```{r heatmaps, fig.height=10, fig.width=12}
res$combined_heatmaps
```

---

## 10. Simulation Validation (True vs. Predicted)

Available only when `true_values` is passed to `fit_and_plot_v22`.  Each panel
shows posterior-mean concentration vs. true concentration on log-log axes.
Coverage = fraction of 95% CIs that contain the true value.

```{r true-vs-pred, fig.height=10, fig.width=12}
if (!is.null(res$combined_true_vs_pred)) {
  res$combined_true_vs_pred
} else {
  cat("(No true_values provided; skipping validation plot)\n")
}
```

---

## 11. Using Real Data

To fit the model to real survey data, replace the simulation steps with your own
data frames and skip `true_values`:

```{r real-data-example, eval=FALSE}
# Load your own data
edna_df <- read.csv("my_edna_reads.csv")   # columns: pcr_replicate_id, lon, lat, species, Nreads
qpcr_df <- read.csv("my_qpcr_data.csv")   # columns: lon, lat, detected, qpcr_log_conc

# PCR amplification efficiencies on log scale (reference species = 0)
# Obtain from standard curves or set to 0 for equal efficiency
alpha_known <- c(0, 0, 0, 0)

# Format
fmt <- format_qpcr_metabar_data_v22(
  edna_data         = edna_df,
  qpcr_data         = qpcr_df,
  alpha_known       = alpha_known,
  reference_species = 1,
  species_names     = c("rockfish_sp1", "rockfish_sp2", "flatfish_sp1", "flatfish_sp2")
)

# Build Stan data — tune rho_prior based on your domain extent
sdat <- make_stan_data_v22(
  fmt,
  rho_prior_mean = 50,    # start at ~half the domain extent
  rho_prior_sd   = 100    # broad; let the data inform
)

# Fit (no true_values for real data)
res <- fit_and_plot_v22(
  fmt,
  stan_file    = "quant_metabar_qpcr_v22.stan",
  stan_data    = sdat,
  chains       = 4,
  iter         = 3000,
  warmup       = 1500,
  adapt_delta  = 0.99,
  save_outputs = TRUE,
  output_dir   = "outputs_real_data"
)
```

---

## 12. Tuning Tips

| Issue | Likely cause | Fix |
|---|---|---|
| Divergences | Difficult posterior geometry | Increase `adapt_delta` (→ 0.995); check `rho` prior vs. domain size |
| Rhat > 1.01 for `rho` | Bimodal or flat posterior | Narrow `rho_prior_sd`; increase `n_locations` |
| Very high nugget fraction | GP too smooth; or over-dispersed eDNA | Check `sigma_nugget_prior_mean`; try lower `log_lambda_mean` |
| Slow sampling | Too many basis functions | Reduce `m_x`, `m_y`; check M < N_locations |
| All-zero qPCR warning | Locations dropped from model | Expected; check that dropped locations are genuine non-detects |
| Poor eDNA proportion fit | Misspecified dispersion priors | Adjust `beta0_prior_sd`, `gamma0_prior_mean` |

**HSGP basis function rule of thumb**: `m_x × m_y` should be 10–20% of
`N_locations` for reasonable approximation.  Error if `m_x * m_y >= N_locations`.

---

## Session Info

```{r session-info}
sessionInfo()
```
